---
title: "Longitudinal protein profiling of blood during childhood into early adulthood"
output: html_document
author: "Sofia Bergström and Simon Kebede Merid"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
editor_options: 
  chunk_output_type: console
---

# Set up
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)

knitr::opts_knit$set(root.dir = "../")
```


## Load packages
```{r}

library(ggpubr) 
library(ggbeeswarm)
library(tidyverse)
require(gridExtra)
library(ggvenn)
library(FactoMineR) 
library(factoextra) 
library(NbClust) 
library(cluster) 
library(patchwork)
library(tidymodels)
library(embed)
library(ggridges) 
library(ggrepel) 
library(readxl)
library(writexl)
library(fpc)
library(doParallel)
library(foreach)
library(corrplot)
library(UpSetR) 
library(clusterProfiler) 
library(org.Hs.eg.db) 
library(grid)
library(png)
library(lme4)     
library(lmerTest) 
library(pbkrtest) 
library(parallel)
library(irr)
library(readr)
mutate <- dplyr::mutate
select <- dplyr::select
rename <- dplyr::rename

```


```{r}
# Selected cutoff
p_value_cutoff <-  0.05
fc_cutoff <- 0.5

# Colors
col_F <-"#ECAA2F"
col_M <- "#9B8FCC"  

col_4 <- "#79A490" 
col_8 <- "#B1DDF1" 
col_16 <- "#8E4C74"
col_24 <- "#1B73B1"

col_protein <-"#D3D3D3"
col_protein_down <-"#FAF3B3" 
col_protein_up <- "#F2E040" 
col_protein_ns <- "#D3D3D3"
col_upset <- "#8F8F8F"

# Prepare for output
if (!dir.exists("output")) {dir.create("output")}
suppdata <- list()
supptables <- list()
```

## Load data
```{r}
# Olink data in long format, one row per sample and assay. Includes daid (sampleid), OlinkID, UniProt, Assay, NPX, LOD, Gene_desc (long gene names), subject_id, follow_up, age and sex
BAMSE_data <- read_csv2("data/data_after_qc/BAMSE_data.csv")|> 
  mutate(subject_id = as.character(subject_id))|> 
  mutate(across(c(sex, follow_up), as.factor))

# Extra clinical data, one row per subject_id. Includes bmi z-scores per follow-up, body fat percentage at follow-up 24, snuff and smoke usage at follow-up 16 and 24, and blood cell counts at follow-up 16 and 24. 
extra_meta_data <- read_csv2("data/clinical/extra_clinical_data.csv")|> 
  mutate(subject_id = as.character(subject_id))
```


# Sample demographics and overview

## Figure 1 

```{r}

# Figure 1A
data_to_plot <- BAMSE_data |>
  select(daid, subject_id, follow_up, age, sex) |> 
  distinct() |> 
  mutate(age = round(age, 1))

p1 <- ggplot(data_to_plot, aes(x = age, y = follow_up, fill = follow_up))+
  geom_density_ridges(color="grey23", alpha=0.8) +
  scale_fill_manual(values=c(col_4, col_8, col_16, col_24))+
  scale_x_continuous(breaks = c(4,8,12, 16,20, 24), minor_breaks = NULL) + 
  theme_minimal()+
  labs(x="Age (years)", y = "Density")+
  theme(legend.position= "none",
        axis.text.y = element_blank())


suppdata$fig1a <- data_to_plot  |> select(follow_up, age) |> arrange(follow_up, age)

# Figure 1B
data_to_plot_prep <- BAMSE_data |>
  select(daid, Assay, NPX, follow_up, sex) |> 
  mutate(sex = case_when(sex == "F" ~ "Female", 
                         sex == "M" ~ "Male") |> factor()) |> 
  pivot_wider(names_from = Assay, values_from = NPX) 

for_colors <- data_to_plot_prep |> 
  select(daid, follow_up, sex)

data_to_plot_scaled <- data_to_plot_prep |> 
  select(-c(daid, follow_up, sex)) |> 
  scale(center = T, scale = T) |> 
  FactoMineR::PCA(graph = F)


p2 <- fviz_pca_ind(data_to_plot_scaled,
             label = "none", 
             habillage = for_colors$follow_up, 
             palette = c(col_4, col_8, col_16, col_24),
             mean.point = FALSE,
             addEllipses = TRUE )+
  labs(title =NULL, color = "Follow-up", fill = "Follow-up", shape = "Follow-up")+
  scale_shape_manual(values = rep(16, 4))+
  theme_light()+
  theme(legend.position = "right", 
        aspect.ratio = 1)

suppdata$fig1b <- data_to_plot_scaled$ind$coord  |> as_tibble() |> select(Dim.1, Dim.2) |> arrange(Dim.1, Dim.2) |> rename(PC1 = Dim.1, PC2 = Dim.2)

# Figure 1C

# Follow-up 4
data_to_plot_prep4 <- data_to_plot_prep |>
  filter(follow_up == 4)

for_colors4 <- data_to_plot_prep4 |> 
  filter(follow_up == 4) |> 
  select(daid, sex)

data_to_plot_scaled <- data_to_plot_prep4 |> 
  select(-c(daid, follow_up, sex)) |> 
  scale(center = T, scale = T)  |>  
  FactoMineR::PCA(graph = F)

p3 <- fviz_pca_ind(data_to_plot_scaled,
             label = "none", 
             habillage = for_colors4$sex, 
             palette = c(col_F, col_M),
             mean.point = FALSE,
             addEllipses = TRUE )+
  labs(title ="Follow-up 4")+
  scale_shape_manual(values = rep(16, 2))+
  theme_light()+
  theme(legend.position = "none")

suppdata$fig1c  <- data_to_plot_scaled$ind$coord  |> as_tibble() |> select(Dim.1, Dim.2) |> arrange(Dim.1, Dim.2) |> mutate(follow_up = 4) |> rename(PC1 = Dim.1, PC2 = Dim.2)

# Follow-up 8
data_to_plot_prep8 <- data_to_plot_prep |>
  filter(follow_up == 8)

for_colors8 <- data_to_plot_prep |> 
  filter(follow_up == 8) |> 
  select(daid, sex)

data_to_plot_scaled <- data_to_plot_prep8 |> 
  select(-c(daid, follow_up, sex)) |> 
  scale(center = T, scale = T)  |>  
  FactoMineR::PCA(graph = F)

p4 <- fviz_pca_ind(data_to_plot_scaled,
             label = "none", 
             habillage = for_colors8$sex, 
             palette = c(col_F, col_M),
             mean.point = FALSE,
             addEllipses = TRUE )+
  labs(title ="Follow-up 8")+
  scale_shape_manual(values = rep(16, 2))+
  theme_light()+
  theme(legend.position = "none")

suppdata$fig1c  <- suppdata$fig1c |> 
  bind_rows(data_to_plot_scaled$ind$coord  |> 
              as_tibble() |> 
              select(Dim.1, Dim.2) |> 
              arrange(Dim.1, Dim.2) |> 
              rename(PC1 = Dim.1, PC2 = Dim.2) |> 
              mutate(follow_up = 8)
            )

# Follow-up 16
data_to_plot_prep16 <- data_to_plot_prep |>
  filter(follow_up == 16)

for_colors16 <- data_to_plot_prep |> 
  filter(follow_up == 16) |> 
  select(daid, sex)

data_to_plot_scaled <- data_to_plot_prep16 |> 
  select(-c(daid, follow_up, sex)) |> 
  scale(center = T, scale = T)  |>  
  FactoMineR::PCA(graph = F)

p5 <- fviz_pca_ind(data_to_plot_scaled,
             label = "none", 
             habillage = for_colors16$sex, 
             palette = c(col_F, col_M),
             mean.point = FALSE,
             addEllipses = TRUE )+
  scale_shape_manual(values = rep(16, 2))+
  labs(title ="Follow-up 16")+
  theme_light()+
  theme(legend.position = "none")

suppdata$fig1c  <- suppdata$fig1c |> 
  bind_rows(data_to_plot_scaled$ind$coord  |> 
              as_tibble() |> 
              select(Dim.1, Dim.2) |> 
              arrange(Dim.1, Dim.2) |> 
              rename(PC1 = Dim.1, PC2 = Dim.2) |> 
              mutate(follow_up = 16)
            )

# Follow-up 24
data_to_plot_prep24 <- data_to_plot_prep |>
  filter(follow_up == 24)

for_colors24 <- data_to_plot_prep |> 
  filter(follow_up == 24) |> 
  select(daid, sex)

data_to_plot_scaled <- data_to_plot_prep24 |> 
  select(-c(daid, follow_up, sex)) |> 
  scale(center = T, scale = T)  |>  
  FactoMineR::PCA(graph = F)

p6 <- fviz_pca_ind(data_to_plot_scaled,
             label = "none", 
             habillage = for_colors24$sex, 
             palette = c(col_F, col_M),
             mean.point = FALSE,
             addEllipses = TRUE )+
  labs(title ="Follow-up 24", color = "Sex", fill = "Sex", shape = "Sex")+
  scale_shape_manual(values = rep(16, 2))+
  theme_light()+
  theme(legend.position = "right")

suppdata$fig1c  <- suppdata$fig1c |> 
  bind_rows(data_to_plot_scaled$ind$coord  |> 
              as_tibble() |> 
              select(Dim.1, Dim.2) |> 
              arrange(Dim.1, Dim.2) |>
              rename(PC1 = Dim.1, PC2 = Dim.2) |> 
              mutate(follow_up = 24)
            )

# Arrange and save Figure 1

p <- ggarrange(
  ggarrange(p1, p2,  nrow = 1, ncol = 2, labels = c("a", "b"), widths = c(1.2,1)),
  ggarrange(p3, p4, p5, p6, nrow = 1, ncol = 4, labels = c("c","", "", ""), widths = c(1,1,1,1.5)), 
  nrow = 2, heights = c(2,1.5))

ggsave(paste0("output/Figure 1 ", Sys.Date(),".pdf"), height = 6, width = 10)

```


# Differences in protein levels between each follow-up

## Figure 2
```{r}

BAMSE_analysis <- BAMSE_data |>
  select(daid, Assay, NPX, subject_id, follow_up, sex) |> 
  distinct() |> 
  mutate(SampleID = daid)


my_theme <-   
  theme_bw()+
  theme(legend.position= "bottom", 
        legend.title = element_blank(), 
        axis.text = element_text(size = 8),   
        axis.title = element_text(size = 10),
        plot.title = element_text(size = 12))

# Figure 2A
data_to_plot <- BAMSE_analysis |> 
  filter(follow_up %in% c(4,8)) |> 
  mutate(follow_up = factor(as.character(follow_up), levels = c(8,4))) |> 
  select(subject_id, Assay, follow_up, NPX) |> 
  pivot_wider(names_from = follow_up, values_from = NPX) |> 
  rename_with(~ paste0("follow_up", .), -c(subject_id, Assay))  |> 
  group_by(Assay) |> 
  summarise(p_value = wilcox.test(follow_up4, follow_up8, paired = TRUE, exact = TRUE)$p.value, 
            estimate = median(follow_up8) - median(follow_up4)) |> 
  ungroup() |> 
  mutate(Adjusted_pval = p.adjust(p_value, method = "BH")) |> 
  arrange(p_value)|> 
  mutate(my_group = case_when(
                              Adjusted_pval < p_value_cutoff & estimate < -fc_cutoff ~ "Down",
                              Adjusted_pval < p_value_cutoff & estimate > fc_cutoff ~ "Up",
                              TRUE ~ "NS") |> factor(levels = c("NS", "Up", "Down")))

age_signif_proteins = data_to_plot |>  
  mutate(comparison = "Follow-up 4 vs 8")

set.seed(42)
p1 <- ggplot(data_to_plot, aes(x= estimate, y= -log10(Adjusted_pval), color = my_group, label=Assay))+
  geom_point(alpha = 0.7, cex = 0.7, shape = 16)+
  geom_text_repel(aes( label= case_when((-log10(Adjusted_pval) > 36) ~ as.character(Assay),
                                        (-log10(Adjusted_pval) > 20)& abs(estimate) > 2.9 ~ as.character(Assay),
                                        (-log10(Adjusted_pval) > 40)& abs(estimate) > 2.6 ~ as.character(Assay),
                                        TRUE ~ "")), size = 1.8, max.overlaps = 15, color = "grey20")+
  scale_color_manual(values=c("NS" = col_protein_ns, "Down" = col_4, "Up" = col_8))+
  labs(y="-log10(p-value)", x = "Log2 Fold Change", title = "Follow-up 4 vs 8")+
  xlim(c(-6.5, 6.5))+
  ylim(c(0,30))+
  my_theme

suppdata$fig2a  <- data_to_plot |> select(Assay, p_value, Adjusted_pval, estimate) |> 
  rename(p_value_raw = p_value, p_value_fdr = Adjusted_pval, Fold_change_log2 = estimate)

# Figure 2B
data_to_plot <- BAMSE_analysis |> 
  filter(follow_up %in% c(8,16)) |> 
  mutate(follow_up = factor(as.character(follow_up), levels = c(16,8))) |> 
  select(subject_id, Assay, follow_up, NPX) |> 
  pivot_wider(names_from = follow_up, values_from = NPX) |> 
  rename_with(~ paste0("follow_up", .), -c(subject_id, Assay))  |> 
  group_by(Assay) |> 
  summarise(p_value = wilcox.test(follow_up8, follow_up16, paired = TRUE, exact = TRUE)$p.value, 
            estimate = median(follow_up16) - median(follow_up8)) |> 
  ungroup() |> 
  mutate(Adjusted_pval = p.adjust(p_value, method = "BH")) |> 
  arrange(p_value)|> 
  mutate(my_group = case_when(
                              Adjusted_pval < p_value_cutoff & estimate < -fc_cutoff ~ "Down",
                              Adjusted_pval < p_value_cutoff & estimate > fc_cutoff ~ "Up",
                              TRUE ~ "NS") |> factor(levels = c("NS", "Up", "Down")))

age_signif_proteins = data_to_plot |>  
  mutate(comparison = "Follow-up 8 vs 16") |> 
  bind_rows(age_signif_proteins)

set.seed(42)
p2 <- ggplot(data_to_plot, aes(x= estimate, y= -log10(Adjusted_pval), color = my_group, label=Assay))+
  geom_point(alpha = 0.7, cex = 0.7, shape = 16)+
  geom_text_repel(aes( label= case_when((-log10(Adjusted_pval) > 27) & estimate < -2.7 ~ as.character(Assay),
                                        (-log10(Adjusted_pval) > 27)& estimate > 2.9 ~ as.character(Assay),
                                        TRUE ~ "")), size = 1.8, max.overlaps = 15, color = "grey20")+
  scale_color_manual(values=c("NS" = col_protein_ns, "Down" = col_8, "Up" = col_16))+
  labs(y="-log10(p-value)", x = "Log2 Fold Change", title = "Follow-up 8 vs 16")+
  xlim(c(-6.5, 6.5))+
  ylim(c(0,30))+
  my_theme

suppdata$fig2b  <- data_to_plot |> select(Assay, p_value, Adjusted_pval, estimate) |> 
  rename(p_value_raw = p_value, p_value_fdr = Adjusted_pval, Fold_change_log2 = estimate)

# Figure 2C
data_to_plot <- BAMSE_analysis |> 
  filter(follow_up %in% c(16,24)) |> 
  mutate(follow_up = factor(as.character(follow_up), levels = c(24,16))) |> 
  select(subject_id, Assay, follow_up, NPX) |> 
  pivot_wider(names_from = follow_up, values_from = NPX) |> 
  rename_with(~ paste0("follow_up", .), -c(subject_id, Assay))  |> 
  group_by(Assay) |> 
  summarise(p_value = wilcox.test(follow_up16, follow_up24, paired = TRUE, exact = TRUE)$p.value, 
            estimate = median(follow_up24) - median(follow_up16)) |> 
  ungroup() |> 
  mutate(Adjusted_pval = p.adjust(p_value, method = "BH")) |> 
  arrange(p_value)|> 
  mutate(my_group = case_when(
                              Adjusted_pval < p_value_cutoff & estimate < -fc_cutoff ~ "Down",
                              Adjusted_pval < p_value_cutoff & estimate > fc_cutoff ~ "Up",
                              TRUE ~ "NS") |> factor(levels = c("NS", "Up", "Down")))


age_signif_proteins = data_to_plot |>  
  mutate(comparison = "Follow-up 16 vs 24") |> 
  bind_rows(age_signif_proteins)

set.seed(42)
p3 <- ggplot(data_to_plot, aes(x= estimate, y= -log10(Adjusted_pval), color = my_group, label=Assay))+
  geom_point(alpha = 0.7, cex = 0.7, shape = 16)+
  geom_text_repel(aes( label= case_when(
                                        (-log10(Adjusted_pval) > 24)& abs(estimate) > 0.6 ~ as.character(Assay),
                                        (-log10(Adjusted_pval) > 20)& estimate > 0 ~ as.character(Assay),
                                        TRUE ~ "")), size = 1.8, max.overlaps = 15, color = "grey20")+
  scale_color_manual(values=c("NS" = col_protein_ns, "Down" = col_16, "Up" = col_24))+
  labs(y="-log10(p-value)", x = "Log2 Fold Change", title = "Follow-up 16 vs 24")+
  xlim(c(-6.5, 6.5))+
  ylim(c(0,30))+
  my_theme

suppdata$fig2c  <- data_to_plot |> select(Assay, p_value, Adjusted_pval, estimate) |> 
  rename(p_value_raw = p_value, p_value_fdr = Adjusted_pval, Fold_change_log2 = estimate)

# Figure 2D
data_to_plot <- age_signif_proteins |> 
  mutate(significant_assay = case_when(my_group %in% c("Up", "Down") ~ 1,
                              my_group == "NS" ~ 0)) |> 
  select(Assay, comparison, significant_assay) |> 
  filter(significant_assay == 1)  |>
  mutate(comparison = case_when(comparison == "Follow-up 4 vs 8" ~ "4 vs 8",
                                comparison == "Follow-up 8 vs 16" ~ "8 vs 16",
                                comparison == "Follow-up 16 vs 24" ~ "16 vs 24")) |> 
  mutate(comparison = factor(comparison, levels = c("4 vs 8", "8 vs 16", "16 vs 24"))) |> 
  group_by(comparison)  |> 
  summarise(elements = list(Assay))  |> 
  deframe()

p4 <- ggvenn(data_to_plot,
       fill_color = c(col_protein_ns, col_protein_ns, col_protein_ns),
       stroke_color = c(col_upset),
       set_name_size = 3,
       text_size = 2.5 ,
       stroke_size = 0.5,
       show_percentage = FALSE )


# Figure 2E
data_to_plot <- age_signif_proteins |> 
  mutate(my_group = case_when(my_group == "NS" ~ "Not significant",
                              my_group == "Down" ~ "Significant down", 
                              my_group == "Up" ~ "Significant up") |> factor(levels = c("Not significant", "Significant up", "Significant down"))) |> 
  mutate(comparison = factor(comparison, levels = c("Follow-up 16 vs 24" , "Follow-up 8 vs 16", "Follow-up 4 vs 8")))  |> 
  group_by(my_group, comparison) |> 
  summarise(n = n())

p5 <- ggplot(data_to_plot, aes(fill=my_group, y=n, x=comparison)) + 
    geom_bar(position="stack", stat="identity")+
    geom_text(aes(label = n), size = 3, hjust = 1.3, vjust = 0.5, position = "stack") +
  scale_fill_manual(values = c(col_protein_ns, col_protein_up, col_protein_down))+
  labs(y = "Number of proteins", x = "", fill = "")+
  theme_light()+
  coord_flip()+
  theme(legend.background=element_blank(), 
        legend.position = "bottom",
        axis.title = element_text(size = 10))

# Figure 2F
selected_proteins <- c("VWC2L",  "AMBN","CD22","SEZ6","CES2", "DSPP", "COL9A1", 
                       "ADM", "CCK", "DMP1", "CCN5", "TRIM25", "MMP3", "CELA2A")

data_to_plot <- BAMSE_data |> 
  select(daid, follow_up, Assay, NPX, Gene_desc) |> 
  filter(Assay %in% selected_proteins) |> 
  mutate(Assay = factor(as.character(Assay), levels = selected_proteins))

p6 <- ggplot(data_to_plot, aes(x = follow_up, y = NPX, fill = follow_up))+
  geom_boxplot(outlier.shape=NA, alpha = 0.5)+
  geom_quasirandom(aes(color = follow_up), width = 0.2, cex = 0.8, alpha = 0.5, shape = 16)+
  labs(x="Follow-up", y = "NPX")+
  scale_fill_manual(values=c(col_4,col_8, col_16, col_24))+
  scale_color_manual(values=c(col_4,col_8, col_16, col_24))+
  facet_wrap(~Assay, ncol = 7, scales = "free")+
  theme_bw()+
  theme(legend.position= "none", legend.title = element_blank(), axis.text = element_text(size = 8), axis.title = element_text(size = 10))

suppdata$fig2f  <- data_to_plot  |> select(Assay, follow_up,  NPX)  |> arrange(Assay, follow_up, NPX)

# Arrange and save Figure 2
p <- ggarrange(
  ggarrange(p1, p2, p3, nrow = 1, ncol = 3, labels = c("a", "b", "c")), 
  ggarrange(p4, p5,  ncol = 2, labels = c("d", "e"), widths = c(0.8, 2.2)),
  p6, nrow = 3, heights = c(1.7,1, 2), labels = c("", "","f"))

ggsave(paste0("output/Figure 2 ", Sys.Date(),".pdf"), height = 11, width = 10)

```

## Supplementary figure 2

```{r}

# Females
BAMSE_F <- BAMSE_data |>
  filter(sex == "F") |> 
  select(subject_id, Assay, follow_up, NPX)

##### 4 vs 8

BAMSE_F_4_8 <- BAMSE_F |> 
  filter(follow_up %in% c(4,8)) |> 
  pivot_wider(names_from = follow_up, values_from = NPX) |> 
  rename_with(~ paste0("follow_up", .), -c(subject_id, Assay))  |> 
  group_by(Assay) |> 
  summarise(p_value = wilcox.test(follow_up4, follow_up8, paired = TRUE, exact = TRUE)$p.value, 
            estimate = median(follow_up8) - median(follow_up4)) |> 
  ungroup() |> 
  mutate(Adjusted_pval = p.adjust(p_value, method = "BH")) |> 
  arrange(p_value)|> 
  mutate(my_group = case_when(
                              Adjusted_pval < p_value_cutoff & estimate < -fc_cutoff ~ "Down",
                              Adjusted_pval < p_value_cutoff & estimate > fc_cutoff ~ "Up",
                              TRUE ~ "NS") |> factor(levels = c("NS", "Up", "Down")))

age_signif_proteins_F = BAMSE_F_4_8 |>  
  mutate(comparison = "Follow-up 4 vs 8")



##### 8 vs 16

BAMSE_F_8_16 <- BAMSE_F |> 
  filter(follow_up %in% c(8,16)) |> 
  pivot_wider(names_from = follow_up, values_from = NPX) |> 
  rename_with(~ paste0("follow_up", .), -c(subject_id, Assay))  |> 
  group_by(Assay) |> 
  summarise(p_value = wilcox.test(follow_up8, follow_up16, paired = TRUE, exact = TRUE)$p.value, 
            estimate = median(follow_up16) - median(follow_up8)) |> 
  ungroup() |> 
  mutate(Adjusted_pval = p.adjust(p_value, method = "BH")) |> 
  arrange(p_value)|> 
  mutate(my_group = case_when(
                              Adjusted_pval < p_value_cutoff & estimate < -fc_cutoff ~ "Down",
                              Adjusted_pval < p_value_cutoff & estimate > fc_cutoff ~ "Up",
                              TRUE ~ "NS") |> factor(levels = c("NS", "Up", "Down")))


age_signif_proteins_F = BAMSE_F_8_16 |>  
  mutate(comparison = "Follow-up 8 vs 16") |> 
  bind_rows(age_signif_proteins_F)

##### 16 vs 24

BAMSE_F_16_24 <- BAMSE_F |> 
  filter(follow_up %in% c(16,24)) |> 
  pivot_wider(names_from = follow_up, values_from = NPX) |> 
  rename_with(~ paste0("follow_up", .), -c(subject_id, Assay))  |> 
  group_by(Assay) |> 
  summarise(p_value = wilcox.test(follow_up16, follow_up24, paired = TRUE, exact = TRUE)$p.value, 
            estimate = median(follow_up24) - median(follow_up16)) |> 
  ungroup() |> 
  mutate(Adjusted_pval = p.adjust(p_value, method = "BH")) |> 
  arrange(p_value)|> 
  mutate(my_group = case_when(
                              Adjusted_pval < p_value_cutoff & estimate < -fc_cutoff ~ "Down",
                              Adjusted_pval < p_value_cutoff & estimate > fc_cutoff ~ "Up",
                              TRUE ~ "NS") |> factor(levels = c("NS", "Up", "Down")))


age_signif_proteins_F = BAMSE_F_16_24 |>  
  mutate(comparison = "Follow-up 16 vs 24") |> 
  bind_rows(age_signif_proteins_F)

# Males
BAMSE_M <-  BAMSE_data |>
  filter(sex == "M") |> 
  select(subject_id, Assay, follow_up, NPX)

##### 4 vs 8

BAMSE_M_4_8 <- BAMSE_M |> 
  filter(follow_up %in% c(4,8)) |> 
  pivot_wider(names_from = follow_up, values_from = NPX) |> 
  rename_with(~ paste0("follow_up", .), -c(subject_id, Assay))  |> 
  group_by(Assay) |> 
  summarise(p_value = wilcox.test(follow_up4, follow_up8, paired = TRUE, exact = TRUE)$p.value, 
            estimate = median(follow_up8) - median(follow_up4)) |> 
  ungroup() |> 
  mutate(Adjusted_pval = p.adjust(p_value, method = "BH")) |> 
  arrange(p_value)|> 
  mutate(my_group = case_when(
                              Adjusted_pval < p_value_cutoff & estimate < -fc_cutoff ~ "Down",
                              Adjusted_pval < p_value_cutoff & estimate > fc_cutoff ~ "Up",
                              TRUE ~ "NS") |> factor(levels = c("NS", "Up", "Down")))

###

age_signif_proteins_M = BAMSE_M_4_8 |>  
  mutate(comparison = "Follow-up 4 vs 8")

##### 8 vs 16

BAMSE_M_8_16 <- BAMSE_M |> 
  filter(follow_up %in% c(8,16)) |>  
  pivot_wider(names_from = follow_up, values_from = NPX) |> 
  rename_with(~ paste0("follow_up", .), -c(subject_id, Assay))  |> 
  group_by(Assay) |> 
  summarise(p_value = wilcox.test(follow_up8, follow_up16, paired = TRUE, exact = TRUE)$p.value, 
            estimate = median(follow_up16) - median(follow_up8)) |> 
  ungroup() |> 
  mutate(Adjusted_pval = p.adjust(p_value, method = "BH")) |> 
  arrange(p_value)|> 
  mutate(my_group = case_when(
                              Adjusted_pval < p_value_cutoff & estimate < -fc_cutoff ~ "Down",
                              Adjusted_pval < p_value_cutoff & estimate > fc_cutoff ~ "Up",
                              TRUE ~ "NS") |> factor(levels = c("NS", "Up", "Down")))


age_signif_proteins_M = BAMSE_M_8_16 |>  
  mutate(comparison = "Follow-up 8 vs 16") |> 
  bind_rows(age_signif_proteins_M)


##### 16 vs 24

BAMSE_M_16_24 <- BAMSE_M |> 
  filter(follow_up %in% c(16,24)) |> 
  pivot_wider(names_from = follow_up, values_from = NPX) |> 
  rename_with(~ paste0("follow_up", .), -c(subject_id, Assay))  |> 
  group_by(Assay) |> 
  summarise(p_value = wilcox.test(follow_up16, follow_up24, paired = TRUE, exact = TRUE)$p.value, 
            estimate = median(follow_up24) - median(follow_up16)) |> 
  ungroup() |> 
  mutate(Adjusted_pval = p.adjust(p_value, method = "BH")) |> 
  arrange(p_value)|> 
  mutate(my_group = case_when(
                              Adjusted_pval < p_value_cutoff & estimate < -fc_cutoff ~ "Down",
                              Adjusted_pval < p_value_cutoff & estimate > fc_cutoff ~ "Up",
                              TRUE ~ "NS") |> factor(levels = c("NS", "Up", "Down")))


age_signif_proteins_M = BAMSE_M_16_24 |>  
  mutate(comparison = "Follow-up 16 vs 24") |> 
  bind_rows(age_signif_proteins_M)

# Combine females and males results and make plot
data_to_plot <-  age_signif_proteins_M |> 
  rename(p_value_M = Adjusted_pval, group_M = my_group, estimate_M = estimate) |> 
  select(Assay, comparison, p_value_M, estimate_M, group_M) |> 
  left_join(age_signif_proteins_F |> 
              rename(p_value_F = Adjusted_pval, group_F = my_group, estimate_F = estimate) |> 
              select(Assay, p_value_F, comparison, group_F, estimate_F), 
            by = c("Assay", "comparison") ) |> 
  mutate(comparison = factor(comparison, levels = c("Follow-up 4 vs 8", "Follow-up 8 vs 16", "Follow-up 16 vs 24"))) |> 
  mutate(my_group = paste0("Males: ", group_M, ", Females: ", group_F)) |> 
  filter(!(my_group == "Males: NS, Females: NS"))

set.seed(193)
p <- ggplot(data_to_plot, aes(x= estimate_M, y= estimate_F, color = my_group, label=Assay))+
  geom_point(alpha = 0.7, cex = 0.7, shape = 16)+
  geom_text_repel(aes( label= case_when(
    (estimate_M < -4) &  (estimate_F < 1) ~ as.character(Assay),
    (estimate_M > 5) ~ as.character(Assay),
    (estimate_F > 2.1) &  (estimate_M < 1) ~ as.character(Assay),
    (estimate_F < 0.5) &  (estimate_M > 3) ~ as.character(Assay),
  TRUE ~ "")), size = 1.8, max.overlaps = 15, color = "grey20")+
  labs(y="Fold change females", x = "Fold change males", title = "")+
  facet_wrap(~comparison)+
  theme_bw()+
  theme(legend.position= "bottom", 
        legend.title = element_blank(), 
        axis.text = element_text(size = 8),   
        axis.title = element_text(size = 10),
        plot.title = element_text(size = 12),
        aspect.ratio = 1)

ggsave(paste0("output/Supplementary figure 2 ", Sys.Date(),".pdf"), height = 4, width = 8)
```

\newpage
# Trajectory analysis to identify different protein clusters

```{r}
## Construct input for clusters

selected_proteins <- age_signif_proteins |> 
  filter(Adjusted_pval < p_value_cutoff & abs(estimate) > fc_cutoff) |> 
  distinct(Assay) |> 
  pull(Assay) 

# Prepare the data
tradjectory_data <- BAMSE_data  |> 
  select(Assay, daid, follow_up, NPX) |> 
  filter(Assay %in% selected_proteins) |> 
  group_by(Assay) |> 
  mutate(scaled_NPX = scale(NPX, center = TRUE, scale = TRUE)) |> 
  ungroup() |> 
  group_by(follow_up, Assay)  |> 
  summarize(median_NPX = median(scaled_NPX, na.rm = TRUE), .groups = "drop") |> 
  pivot_wider(names_from = follow_up, values_from = median_NPX)  |> 
  rename_with(~ paste0("follow_up", .), -Assay)  |> 
  column_to_rownames(var = "Assay") |> 
  as.data.frame()


set.seed(2008)
dist_matrix <- tradjectory_data |> 
  as.matrix() |> 
  dist(method = "euclidean")

```


## Supplementary figure 1
```{r}
my_theme <-   
  theme_bw()+
  theme(axis.text = element_text(size = 8),   
        axis.title = element_text(size = 10),
        plot.title = element_text(size = 12))

p1 <- fviz_nbclust(tradjectory_data |>
as.matrix(), FUN = hcut, method = "silhouette")+
    labs(
    title = "",
    x = "Number of clusters (k)")+
  my_theme


p2 <- fviz_nbclust(tradjectory_data |>
   as.matrix(), FUN = hcut, method = "gap_stat")+
    labs(
    title = "",
    x = "Number of clusters (k)")+
  my_theme

set.seed(2008)

# Indices to plot
plot_indices <- c("cindex")

# Run NbClust
nb_list <- lapply(plot_indices, function(idx) {
  NbClust(
    tradjectory_data, diss = NULL,
    distance = "euclidean",
    min.nc = 2,
    max.nc = 10,
    method = "ward.D2",
    index = idx, alphaBeale = 0.1
  )
})
names(nb_list) <- plot_indices

# Extract index values
index_df <- bind_rows(lapply(names(nb_list), function(idx) {
  data.frame(
    k = as.numeric(names(nb_list[[idx]]$All.index)),
    value = as.numeric(nb_list[[idx]]$All.index),
    index = idx
  )
}))|>  mutate(index = as.character(index))

p3 <- ggplot(index_df, aes(x = k, y = value)) +
  geom_line(linewidth = 0.5, color = "#4682b4") +
  geom_point(size = 1.8, color = "#4682b4") +
  geom_vline(
    aes(xintercept = 8),
    linetype = "dashed",
    color = "steelblue",
    linewidth = 0.5
  )  +
  scale_x_continuous(breaks = 2:10) +
  theme_classic()+
  labs(
    title = "",
    x = "Number of clusters (k)",
    y = "C-index") + # (lower is better)
  my_theme


# Combine the plots
p <- ggarrange(p1, p2, p3, nrow = 1, ncol = 3, labels = c("a", "b", "c"), widths = c(1,1,1))

ggsave(paste0("output/Supplementary figure 1 ", Sys.Date(),".pdf"), height = 5, width = 10)

```


```{r}
# Evaluate cluster stability 

#Define cluster function wrapper
wardD2_fun <- function(d, k) {
  hclustCBI(as.matrix(d), 
            method = "ward.D2", 
            cut = "number", 
            k = k)
}

# Setup parallel 
nc <- max(1, parallel::detectCores() - 2)  
cl <- makeCluster(nc)
registerDoParallel(cl)

# Set bootstrap parameters
B_total <- 100                    
B_per_worker <- ceiling(B_total / nc)  

# Run clusterboot in parallel
set.seed(2008)
results <- foreach(i = 1:nc, .packages = "fpc") %dopar% {
  clusterboot(
    dist_matrix,
    B = B_per_worker,
    distances = TRUE,        
    bootmethod = "boot",
    clustermethod = wardD2_fun,
    k = 8                   
  )
}

stopCluster(cl)

# Aggregate results
bootmean_mat <- do.call(rbind, lapply(results, `[[`, "bootmean"))
bootbrd_mat  <- do.call(rbind, lapply(results, `[[`, "bootbrd"))

bootmean_avg <- colMeans(bootmean_mat, na.rm = TRUE)  # average Jaccard per cluster
bootbrd_sum  <- colSums(bootbrd_mat, na.rm = TRUE)    # sum of cluster dissolutions

# Use partition from first run as representative
final_partition <- results[[1]]$result$partition
clusters_from_stability <- tibble(Assay = names(final_partition), Cluster = final_partition)

# Add cluster labels to the data
tradjectory_data$Assay <-rownames(tradjectory_data)
tradjectory_data <- tradjectory_data |> 
  left_join(clusters_from_stability, by = "Assay")

saved_protein_cluster <- tradjectory_data |> 
  distinct(Assay, Cluster) |> 
  as_tibble()

```

## Figure 3
```{r}
# Figure 3A
cluster_sizes <- saved_protein_cluster  |> 
  group_by(Cluster)  |> 
  summarise(Size = n()) |> 
  mutate(Cluster_Label = paste0("Cluster ", Cluster, "\n", Size, " proteins")) |> 
  select(Cluster, Cluster_Label)

data_to_plot <- tradjectory_data  |> 
  pivot_longer(cols = starts_with("follow_up"), 
               names_to = "follow_up", 
               values_to = "NPX")  |> 
  mutate(follow_up = case_when(follow_up == "follow_up4" ~ "4",
                               follow_up == "follow_up8" ~ "8",
                               follow_up == "follow_up16" ~ "16",
                               follow_up == "follow_up24" ~ "24")) |> 
  mutate(follow_up = factor(follow_up, levels = c("4", "8", "16", "24"))) |> 
  left_join(cluster_sizes, by = "Cluster") |> 
  mutate(Cluster = paste0("Cluster ", Cluster))

p1 <- 
  ggplot(data_to_plot, aes(x = follow_up, y = NPX, group = Assay)) + 
  geom_line(color = col_protein, alpha = 0.2) +  
  geom_smooth(aes(group = 1), method = "loess", span = 0.7,
              se = FALSE, color = "grey30", linewidth = 1, linetype = "solid") +
  facet_wrap(~ Cluster_Label, ncol = 8) +  
  ylim(-1.7,1.7)+
  labs(
       x = "Follow-up",
       y = "Scaled NPX") + 
  theme_bw() +
  theme(legend.position = "bottom") 


suppdata$fig3a  <- data_to_plot |> select(Cluster, Assay, follow_up, NPX) |> 
  pivot_wider(names_from = follow_up, values_from = NPX, names_prefix = "follow_up_") |> 
  arrange(Cluster, Assay)

# Figure 3B
data_to_plot <- age_signif_proteins |> 
  mutate(my_group = case_when(Adjusted_pval > p_value_cutoff ~ "Not significant",
                              Adjusted_pval < p_value_cutoff & estimate < -fc_cutoff ~ "Significant down", 
                              Adjusted_pval < p_value_cutoff & estimate > fc_cutoff ~ "Significant up", TRUE ~"Not significant") |> factor(levels = c("Not significant", "Significant up", "Significant down"))) |> 
  mutate(comparison = case_when(comparison == "Follow-up 16 vs 24" ~ "16 vs 24",
                                comparison == "Follow-up 8 vs 16" ~ "8 vs 16",
                                comparison == "Follow-up 4 vs 8" ~ "4 vs 8")) |>
    mutate(comparison = factor(comparison, levels = c("4 vs 8", "8 vs 16", "16 vs 24"))) |> 
  left_join(saved_protein_cluster, by = "Assay")|> 
  mutate(Cluster = paste0("Cluster ", Cluster)) |> 
  filter(Cluster != "Cluster NA") |> 
  group_by(my_group, comparison, Cluster) |> 
  summarise(n = n()) |> 
  mutate(text_for_plot = case_when(n < 45 ~ NA,  TRUE ~ n))


p2 <- ggplot(data_to_plot, aes(fill=my_group, y=n, x=comparison)) + 
    geom_bar(position="stack", stat="identity")+
    geom_text(aes(label = text_for_plot), size = 2.5, hjust = 0.5, vjust = 1.3, position = "stack") +
  scale_fill_manual(values = c(col_protein_ns, col_protein_up, col_protein_down))+
  labs(y = "Number of proteins", x = "", fill = "")+
  theme_bw()+
  ggh4x::facet_wrap2(~ Cluster, , ncol = 8, scales = "fixed", axes = "all")+
  theme(legend.background=element_blank(), legend.position = "bottom", axis.text.x = element_text(size = 9, angle = 45, hjust = 1))


# Figure 3C
selected_proteins <- c("CEP250","UBR2", "APOA2", "SEPTIN7", "SPINT3", "VRK1", "KLK14", "IGFBP1")

data_to_plot <- BAMSE_data |> 
  select(daid, subject_id, follow_up, Assay, NPX) |> 
  left_join(saved_protein_cluster, by = "Assay") |> 
  filter(Assay %in% selected_proteins) |> 
  mutate(Assay = factor(Assay, levels = selected_proteins)) |> 
  mutate(Cluster_assay = paste0("Cluster ", Cluster, "\n", Assay))

p3 <- 
  ggplot(data_to_plot, aes(x = follow_up, y = NPX, group = subject_id))+
  geom_line(color = col_protein, alpha = 0.4) +  
  geom_smooth(aes(group = 1), method = "loess", span = 0.7, 
              se = FALSE, color = "grey30", linewidth = 1, linetype = "solid") +
  labs(x="Follow up", y = "NPX")+
  facet_wrap(~ Cluster_assay, scales = "free_y", ncol = 8) +
  theme_bw()


suppdata$fig3c  <- data_to_plot |> select(Assay, Cluster, subject_id, follow_up, follow_up, NPX) |> pivot_wider(names_from = follow_up, values_from = NPX, names_prefix = "Follow_up_") |>
  relocate(Assay, Cluster, subject_id, Follow_up_4, Follow_up_8, Follow_up_16, Follow_up_24) |>
  select(-subject_id) |> 
  arrange(Cluster, Follow_up_4, Follow_up_8, Follow_up_16, Follow_up_24) 
  

# Arrange and save Figure 3

p <-  ggarrange(p1, p2, p3, nrow = 3, ncol = 1, labels = c("a", "b", "c"), heights = c(1,1.2,1))

ggsave(paste0("output/Figure 3 ", Sys.Date(),".pdf"), height = 10, width = 10)

```
\newpage

## Table 2 
```{r}

safe_enrichGO <- possibly(enrichGO, otherwise = NULL)

universe_proteins <- saved_protein_cluster |> 
  left_join(BAMSE_data |> select(Assay, UniProt) |> distinct(), by = "Assay") |> 
  pull(UniProt) |> 
  unique()

enrichment_result_tradjectories <- saved_protein_cluster |> 
  left_join(BAMSE_data |> select(Assay, UniProt) |> distinct(), by = "Assay") |> 
  group_by(Cluster) |> 
  summarise(
    ego = list(
      safe_enrichGO(
        gene = UniProt,
        universe = universe_proteins,
        OrgDb = org.Hs.eg.db,
        keyType = "UNIPROT",  
        ont = "BP",
        pAdjustMethod = "BH",
        pvalueCutoff = 0.05,
        qvalueCutoff = 0.2,
        readable = TRUE
      )
    ),
    .groups = "drop"
  )


###
enrichment_result_tradjectories_table <- enrichment_result_tradjectories  |> 
  mutate(ego_df = map(ego, ~ as_tibble(.x@result))) |> 
  select(Cluster, ego_df)  |> 
  unnest(ego_df) |> 
  filter(p.adjust < 0.05) |> 
  arrange(Cluster, pvalue) |> 
  select(Cluster, ID, Description, p.adjust, pvalue, Count, geneID) |> 
  mutate(p.adjust = format(p.adjust, scientific = TRUE, digits = 2, nsmall = 2)) |> 
  mutate(pvalue = format(pvalue, scientific = TRUE, digits = 2, nsmall = 2)) 


supptable2  <- enrichment_result_tradjectories_table |>
  rename("p-value raw" = pvalue, "p-value FDR" = p.adjust)

enrichment_result_tradjectories_table |> 
  group_by(Cluster)  |>
  slice_min(order_by = pvalue, n = 6)  |>
  ungroup() |>
  arrange(Cluster, pvalue) |> 
  select(-c(geneID)) |> 
  rename("p-value raw" = pvalue, "p-value FDR" = p.adjust, "Number of proteins"  = Count) |> 
  write.csv(paste0("output/Table 2 ", Sys.Date(), ".csv"))

```

\newpage
# Sex-associated protein profiles
## Figure 4
```{r}

BAMSE_analysis <- BAMSE_data |>
  select(daid, Assay, NPX, subject_id, follow_up, sex) |>
  distinct() |> 
  mutate(SampleID = daid) |> 
    mutate(sex = case_when(sex == "F" ~ "Female", 
                         sex == "M" ~ "Male") |> factor())

my_theme <-   theme_bw()+
  theme(legend.position= "none", 
        legend.title = element_blank(), 
        axis.text = element_text(size = 8),   
        axis.title = element_text(size = 10),
        plot.title = element_text(size = 12))

# Figure 4A
estimate4 <- BAMSE_analysis |> 
  filter(follow_up == 4) |> 
  select(subject_id, Assay, NPX, sex) |> 
  group_by(Assay, sex) |> 
  summarise(median = median(NPX)) |> 
  ungroup() |> 
  pivot_wider(names_from = sex, values_from = median) |> 
  mutate(estimate = Female - Male) |> 
  select(Assay, estimate)

data_to_plot4 <- BAMSE_analysis |> 
  filter(follow_up == 4) |> 
  select(subject_id, Assay, NPX, sex) |> 
  group_by(Assay) |> 
  summarise(p_value = wilcox.test(NPX ~ sex, exact = TRUE)$p.value) |> 
  ungroup() |> 
  left_join(estimate4, by = "Assay") |> 
  mutate(Adjusted_pval = p.adjust(p_value, method = "BH")) |> 
  arrange(p_value)|> 
  mutate(my_group = case_when(
                              Adjusted_pval < p_value_cutoff & estimate < -fc_cutoff ~ "Down",
                              Adjusted_pval < p_value_cutoff & estimate > fc_cutoff ~ "Up",
                              TRUE ~ "NS") |> factor(levels = c("NS", "Up", "Down")))

rm(estimate4)

set.seed(42) 
p1 <- ggplot(data_to_plot4, aes(x= estimate, y= -log10(Adjusted_pval), color = my_group, label=Assay))+
  geom_point(alpha = 0.7, cex = 0.7)+
  geom_text_repel(aes( label= case_when((-log10(p_value) > 17) ~ as.character(Assay),(-log10(Adjusted_pval) > 3) ~ as.character(Assay),
                                        (estimate > 2.3) ~ as.character(Assay)
                                        ,  TRUE ~ "")), size = 2, max.overlaps = 10, color = "grey20")+
    scale_color_manual(values=c("NS" = col_protein_ns, "Down" = col_M, "Up" = col_F))+
  labs(y="-log10(p-value)", x = "Log2 Fold Change", title = "Follow-up 4")+
  xlim(c(-7.8, 7.8))+
  ylim(c(0,30))+
  my_theme

suppdata$fig4a  <- data_to_plot4 |> select(Assay, p_value, Adjusted_pval, estimate) |> 
  rename(p_value_raw = p_value, p_value_fdr = Adjusted_pval, Fold_change_log2 = estimate)


# Figure 4B
estimate8 <- BAMSE_analysis |> 
  filter(follow_up == 8) |> 
  select(subject_id, Assay, NPX, sex) |> 
  group_by(Assay, sex) |> 
  summarise(median = median(NPX)) |> 
  ungroup() |> 
  pivot_wider(names_from = sex, values_from = median) |> 
  mutate(estimate = Female - Male) |> 
  select(Assay, estimate)

data_to_plot8 <- BAMSE_analysis |> 
  filter(follow_up == 8) |> 
  select(subject_id, Assay, NPX, sex) |> 
  group_by(Assay) |> 
  summarise(p_value = wilcox.test(NPX ~ sex, exact = TRUE)$p.value) |> 
  ungroup() |> 
  left_join(estimate8, by = "Assay") |> 
  mutate(Adjusted_pval = p.adjust(p_value, method = "BH")) |> 
  arrange(p_value)|> 
  mutate(my_group = case_when(
                              Adjusted_pval < p_value_cutoff & estimate < -fc_cutoff ~ "Down",
                              Adjusted_pval < p_value_cutoff & estimate > fc_cutoff ~ "Up",
                              TRUE ~ "NS") |> factor(levels = c("NS", "Up", "Down")))

rm(estimate8)

set.seed(42)
p2 <- ggplot(data_to_plot8, aes(x= estimate, y= -log10(Adjusted_pval), color = my_group, label=Assay))+
  geom_point(alpha = 0.7, cex = 0.7)+
  geom_text_repel(aes( label= case_when((-log10(p_value) > 17) ~ as.character(Assay), (-log10(Adjusted_pval) > 3) ~ as.character(Assay),
                                        (estimate > 2.3) ~ as.character(Assay)
                                        ,  TRUE ~ "")), size = 2, max.overlaps = 10, color = "grey20")+
    scale_color_manual(values=c("NS" = col_protein_ns, "Down" = col_M, "Up" = col_F))+
  labs(y="-log10(p-value)", x = "Log2 Fold Change", title = "Follow-up 8")+
  xlim(c(-7.8, 7.8))+
  ylim(c(0,30))+
  my_theme

suppdata$fig4b  <- data_to_plot8 |> select(Assay, p_value, Adjusted_pval, estimate) |> 
  rename(p_value_raw = p_value, p_value_fdr = Adjusted_pval, Fold_change_log2 = estimate)

# Figure 4C
estimate16 <- BAMSE_analysis |> 
  filter(follow_up == 16) |> 
  select(subject_id, Assay, NPX, sex) |> 
  group_by(Assay, sex) |> 
  summarise(median = median(NPX)) |> 
  ungroup() |> 
  pivot_wider(names_from = sex, values_from = median) |> 
  mutate(estimate = Female - Male) |> 
  select(Assay, estimate)

data_to_plot16 <- BAMSE_analysis |> 
  filter(follow_up == 16) |> 
  select(subject_id, Assay, NPX, sex) |> 
  group_by(Assay) |> 
  summarise(p_value = wilcox.test(NPX ~ sex, exact = TRUE)$p.value) |> 
  ungroup() |> 
  left_join(estimate16, by = "Assay") |> 
  mutate(Adjusted_pval = p.adjust(p_value, method = "BH")) |> 
  arrange(p_value)|> 
  mutate(my_group = case_when(
                              Adjusted_pval < p_value_cutoff & estimate < -fc_cutoff ~ "Down",
                              Adjusted_pval < p_value_cutoff & estimate > fc_cutoff ~ "Up",
                              TRUE ~ "NS") |> factor(levels = c("NS", "Up", "Down")))

rm(estimate16)

set.seed(42)
p3 <- ggplot(data_to_plot16, aes(x= estimate, y= -log10(Adjusted_pval), color = my_group, label=Assay))+
  geom_point(alpha = 0.7, cex = 0.7)+
  geom_text_repel(aes( label= case_when((-log10(Adjusted_pval) > 17) ~ as.character(Assay),
                                        (-log10(Adjusted_pval) > 5) & (estimate > 2.3) ~ as.character(Assay)
                                        ,  TRUE ~ "")), size = 2, max.overlaps = 10, color = "grey20")+
    scale_color_manual(values=c("NS" = col_protein_ns, "Down" = col_M, "Up" = col_F))+
  labs(y="-log10(p-value)", x = "Log2 Fold Change", title = "Follow-up 16")+
  xlim(c(-7.8, 7.8))+
  ylim(c(0,30))+
  my_theme


suppdata$fig4c  <- data_to_plot16 |> select(Assay, p_value, Adjusted_pval, estimate) |> 
  rename(p_value_raw = p_value, p_value_fdr = Adjusted_pval, Fold_change_log2 = estimate)

# Figure 4D
estimate24 <- BAMSE_analysis |> 
  filter(follow_up == 24) |> 
  select(subject_id, Assay, NPX, sex) |> 
  group_by(Assay, sex) |> 
  summarise(median = median(NPX)) |> 
  ungroup() |> 
  pivot_wider(names_from = sex, values_from = median) |> 
  mutate(estimate = Female - Male) |> 
  select(Assay, estimate)

data_to_plot24 <- BAMSE_analysis |> 
  filter(follow_up == 24) |> 
  select(subject_id, Assay, NPX, sex) |> 
  group_by(Assay) |> 
  summarise(p_value = wilcox.test(NPX ~ sex, exact = TRUE)$p.value) |> 
  ungroup() |> 
  left_join(estimate24, by = "Assay") |> 
  mutate(Adjusted_pval = p.adjust(p_value, method = "BH")) |> 
  arrange(p_value)|> 
  mutate(my_group = case_when(
                              Adjusted_pval < p_value_cutoff & estimate < -fc_cutoff ~ "Down",
                              Adjusted_pval < p_value_cutoff & estimate > fc_cutoff ~ "Up",
                              TRUE ~ "NS") |> factor(levels = c("NS", "Up", "Down")))

rm(estimate24)

set.seed(42)
p4 <- ggplot(data_to_plot24, aes(x= estimate, y= -log10(Adjusted_pval), color = my_group, label=Assay))+
  geom_point(alpha = 0.7, cex = 0.7)+
  geom_text_repel(aes( label= case_when((-log10(p_value) > 17) ~ as.character(Assay),
                                        (estimate > 2.5) ~ as.character(Assay)
                                        ,  TRUE ~ "")), size = 2, max.overlaps = 10, color = "grey20")+
    scale_color_manual(values=c("NS" = col_protein_ns, "Down" = col_M, "Up" = col_F))+
  labs(y="-log10(p-value)", x = "Log2 Fold Change", title = "Follow-up 24")+
  xlim(c(-7.8, 7.8))+
  ylim(c(0,30))+
  my_theme


suppdata$fig4d  <- data_to_plot24 |> select(Assay, p_value, Adjusted_pval, estimate) |> 
  rename(p_value_raw = p_value, p_value_fdr = Adjusted_pval, Fold_change_log2 = estimate)

# Save p-values
save_p_values_sex <- data_to_plot4 |> 
  select(Assay, estimate, Adjusted_pval, my_group) |> 
  mutate(comparison = "Follow up 4") |>
  bind_rows(data_to_plot8 |> 
              select(Assay, estimate, Adjusted_pval, my_group) |> 
              mutate(comparison = "Follow up 8")) |> 
  bind_rows(data_to_plot16 |> 
              select(Assay, estimate, Adjusted_pval, my_group) |> 
              mutate(comparison = "Follow up 16")) |> 
  bind_rows(data_to_plot24 |> 
              select(Assay, estimate, Adjusted_pval, my_group) |> 
              mutate(comparison = "Follow up 24"))

save_p_values_sex_including_raw_p <- data_to_plot4 |> 
  select(Assay, estimate, p_value, Adjusted_pval, my_group) |> 
  mutate(comparison = "Follow up 4") |>
  bind_rows(data_to_plot8 |> 
              select(Assay, estimate, p_value, Adjusted_pval, my_group) |> 
              mutate(comparison = "Follow up 8")) |> 
  bind_rows(data_to_plot16 |> 
              select(Assay, estimate, p_value, Adjusted_pval, my_group) |> 
              mutate(comparison = "Follow up 16")) |> 
  bind_rows(data_to_plot24 |> 
              select(Assay, estimate, p_value, Adjusted_pval, my_group) |> 
              mutate(comparison = "Follow up 24"))


# Figure 4E
data_to_plot <- data_to_plot4 |> 
  mutate(`4` = case_when(my_group %in% c("Up", "Down") ~1, TRUE~0)) |> 
  select(Assay, `4`) |> 
  left_join(data_to_plot8 |> 
              mutate(`8` = case_when(my_group %in% c("Up", "Down")~1, TRUE~0)) |> 
              select(Assay, `8`), by = "Assay") |> 
  left_join(data_to_plot16 |> 
              mutate(`16` = case_when(my_group %in% c("Up", "Down")~1, TRUE~0)) |> 
              select(Assay, `16`), by = "Assay") |> 
  left_join(data_to_plot24 |> 
              mutate(`24` = case_when(my_group %in% c("Up", "Down")~1, TRUE~0)) |> 
              select(Assay, `24`), by = "Assay") 

sex_binary_results  <- data_to_plot |> 
  rename(follow_up_4 = `4`, follow_up_8 = `8`, follow_up_16 = `16`, follow_up_24 = `24`)
  
p6 <- upset(data_to_plot |> 
        select(-Assay) |> 
        as.data.frame(), 
      order.by = "freq", 
      keep.order = TRUE, 
      sets = c("24", "16", "8", "4"),
      mb.ratio = c(0.6, 0.4),
      point.size = 2.8,
      main.bar.color = col_upset,
      matrix.color = col_upset,
      sets.bar.color = col_upset 
      )

# Save the UpSetR plot to a temporary PNG file
tmp_file <- tempfile(fileext = ".png")
png(tmp_file, width = 800, height = 600, res = 200)
dev.off()

# Read the PNG file and convert it into a grid object
p6_grob <- rasterGrob(readPNG(tmp_file), interpolate = TRUE)

# Figure 4F
selected_proteins <- data_to_plot |> 
  filter(`16` == 1 & `24` == 1) |> 
  pull(Assay)

data_to_plot_prep <- BAMSE_data |>
  select(daid, Assay, NPX, follow_up, sex) |> 
  filter(follow_up %in% c(24)) |> 
  filter(Assay %in% selected_proteins) |> 
  pivot_wider(names_from = Assay, values_from = NPX) 

for_colors <- data_to_plot_prep |> 
  select(daid, sex)

data_to_plot_scaled <- data_to_plot_prep |> 
  select(-c(daid, follow_up, sex)) |> 
  scale(center = T, scale = T)  |>  
  FactoMineR::PCA(graph = F)

p7 <- fviz_pca_ind(data_to_plot_scaled,
             label = "none", 
             habillage = for_colors$sex, 
             palette = c(col_F, col_M),
             addEllipses = TRUE )+
  labs(title ="")+
  scale_shape_manual(values = rep(16, 2))+
  theme_light()+
  theme(legend.position = "none", 
        legend.key.size = unit(0.3, "cm"))

suppdata$fig4f <- data_to_plot_scaled$ind$coord  |> as_tibble() |> select(Dim.1, Dim.2) |> arrange(Dim.1, Dim.2) |> rename(PC1 = Dim.1, PC2 = Dim.2)


# Figure 4G
selected_proteins = c("ZP4","XG","PAEP", "LEP","OVGP1" ,"GH1", "SHBG",  "AMELY", "SPESP1", "TXLNGY", "SSC4D", "ACAN","INSL3", "SPINT3")

data_to_plot <- BAMSE_data |>
  filter(Assay %in% selected_proteins) |>
  mutate(Assay = factor(as.character(Assay), levels = selected_proteins)) |> 
  select(daid, Assay, NPX, subject_id, follow_up, sex, Gene_desc) |>  
  distinct()

p5 <- 
  ggplot(data_to_plot, aes(x = follow_up, y = NPX, fill = sex))+
  geom_boxplot(outlier.shape=NA, alpha = 0.5)+
  geom_quasirandom(aes(color = sex, group = sex), width = 0.1, alpha = 0.7, cex = 0.8, dodge.width = 0.75, shape = 16 )+
  labs(x="", y = "NPX")+
  scale_fill_manual(values=c(col_F,col_M))+
  scale_color_manual(values=c(col_F,col_M))+
  facet_wrap(~Assay, ncol = 7, scales = "free")+
  labs(x = "Follow-up")+
  theme_bw()+
  theme(legend.position = "bottom",
        legend.title = element_blank(), 
        axis.text = element_text(size = 8))

suppdata$fig4g  <- data_to_plot  |> select(Assay, follow_up,  NPX)  |> arrange(Assay, follow_up, NPX)

# Arrange and save Figure 4

p <- ggarrange(
  ggarrange(p1, p2, p3, p4, nrow = 1, ncol = 4, labels = c("a", "b", "c", "d")),
  ggarrange(p6_grob, p7, ncol = 2, widths = c(1,1), labels = c("e", "f")), 
  ggarrange(p5, nrow = 1, labels = c("g")), 
  nrow = 3, heights = c(1, 1.2, 2))

ggsave(paste0("output/Figure 4 ", Sys.Date(),".pdf"), height = 11, width = 10)

```

## Table 3
```{r}

safe_enrichGO <- possibly(enrichGO, otherwise = NULL)

universe_proteins <- BAMSE_data |> 
  distinct(Assay, UniProt) |>  
  pull(UniProt) |> 
  unique()

selected_proteins1 <- sex_binary_results |> 
  filter(follow_up_16 == 1) |> 
  select(Assay) |> 
  mutate(my_group = "protein_16")

selected_proteins2 <- sex_binary_results |> 
  filter(follow_up_24 == 1) |> 
  select(Assay) |> 
  mutate(my_group = "protein_24")

selected_proteins3 <- sex_binary_results |> 
  filter(follow_up_16 == 1 & follow_up_24 == 1) |> 
  select(Assay) |> 
  mutate(my_group = "protein_16_24")

selected_proteins <- bind_rows(selected_proteins1, selected_proteins2, selected_proteins3) |> 
  left_join(BAMSE_data |> distinct(Assay, UniProt), by = "Assay")

rm(selected_proteins1, selected_proteins2, selected_proteins3)

enrichment_result_sex <- selected_proteins |> 
  group_by(my_group) |> 
  summarise(
    ego = list(
      safe_enrichGO(
        gene = UniProt,
        universe = universe_proteins,
        OrgDb = org.Hs.eg.db,
        keyType = "UNIPROT",  
        ont = "BP",
        pAdjustMethod = "BH",
        pvalueCutoff = 0.05,
        qvalueCutoff = 0.2,
        readable = TRUE
      )
    ),
    .groups = "drop"
  )


###
enrichment_result_sex  |> 
  mutate(ego_df = map(ego, ~ as_tibble(.x@result))) |> 
  select(my_group, ego_df)  |> 
  unnest(ego_df) |> 
  group_by(my_group)  |> 
  slice_min(order_by = pvalue, n = 10)  |> 
  ungroup() |> 
  arrange(my_group, pvalue) |> 
  select(my_group, ID, Description, pvalue, p.adjust, Count) |> 
  mutate(p.adjust = format(p.adjust, scientific = TRUE, digits = 2, nsmall = 2)) |> 
  mutate(pvalue = format(pvalue, scientific = TRUE, digits = 2, nsmall = 2)) |> 
  rename("p-value raw" = pvalue, "p-value FDR" = p.adjust, "Number of proteins" = Count, "Protein set" = my_group) |> 
  write.csv2(paste0("output/Table 3 ", Sys.Date(), ".csv"))


## Supp info

supptable5  <- enrichment_result_sex  |> 
  mutate(ego_df = map(ego, ~ as_tibble(.x@result))) |> 
  select(my_group, ego_df)  |> 
  unnest(ego_df) |> 
  filter(pvalue < 0.05) |> 
  arrange(my_group, pvalue) |> 
  select(my_group, ID, Description, pvalue, p.adjust, Count, geneID) |> 
  mutate(p.adjust = format(p.adjust, scientific = TRUE, digits = 2, nsmall = 2)) |> 
  mutate(pvalue = format(pvalue, scientific = TRUE, digits = 2, nsmall = 2)) |> 
  rename("p-value raw" = pvalue, "p-value FDR" = p.adjust, "Number of proteins" = Count, "Protein set" = my_group) 

```

\newpage

# Sex-related-differences in proteins after adjustment for covariates

```{r}
proteins_16_24 <- sex_binary_results |> 
  filter(follow_up_16 == 1 & follow_up_24 == 1) |> 
  pull(Assay)

proteins_16 <- sex_binary_results |> 
  filter(follow_up_16 == 1) |> 
  pull(Assay)

proteins_24 <- sex_binary_results |> 
  filter(follow_up_24 == 1) |> 
  pull(Assay)

data_16 <- BAMSE_data |>
  select(daid, subject_id, Assay, NPX, follow_up, sex) |> 
  filter(follow_up %in% c(16)) |>  
  left_join(extra_meta_data |> select(subject_id, snuff_16, smoke_16, bmi_z_score_16y, erytrocyt_16, leukocyt_16), by = "subject_id") |> 
  rename(snuff = snuff_16, smoke = smoke_16, bmi = bmi_z_score_16y, erytrocyt = erytrocyt_16, leukocyt = leukocyt_16) |> 
  mutate(snuff = factor(snuff), smoke = factor(smoke))

data_24 <- BAMSE_data |>
  select(daid, subject_id, Assay, NPX, follow_up, sex) |> 
  filter(follow_up %in% c(24)) |> 
  left_join(extra_meta_data |> select(subject_id, snuff_24, smoke_24, bmi_z_score_24y, bioimp_body_fat, erytrocyt_24, leukocyt_24), by = "subject_id") |> 
  rename(snuff = snuff_24, smoke = smoke_24, bmi = bmi_z_score_24y, erytrocyt = erytrocyt_24, leukocyt = leukocyt_24) |> 
  mutate(snuff = factor(snuff), smoke = factor(smoke))

extract_number_significant <- function(my_data, my_proteins, my_model) {
    my_data  |>
    filter(Assay %in% my_proteins) |> 
    group_by(Assay)  |>  
    do(tidy(lm(my_model, data = .))) |> 
    filter(term == "sexM") |> 
    filter(p.value < p_value_cutoff) |> 
    pull(Assay) |> 
    length()
}

model_1 <- NPX ~ sex + bmi + snuff + smoke
model_2 <- NPX ~ sex + bioimp_body_fat + snuff + smoke
model_3 <- NPX ~ sex + erytrocyt + leukocyt

extract_number_significant(data_16, proteins_16_24, model_1)
extract_number_significant(data_24, proteins_16_24, model_1)
extract_number_significant(data_16, proteins_16, model_1)
extract_number_significant(data_24, proteins_24, model_1)

extract_number_significant(data_24, proteins_24, model_2)

extract_number_significant(data_16, proteins_16_24, model_3)
extract_number_significant(data_24, proteins_16_24, model_3)
extract_number_significant(data_16, proteins_16, model_3)
extract_number_significant(data_24, proteins_24, model_3)

# Extract proteins
extract_significant_assays_sex <- function(my_data, my_proteins, my_model) {
    my_data  |>
    filter(Assay %in% my_proteins) |> 
    group_by(Assay)  |>  
    do(tidy(lm(my_model, data = .))) |> 
    filter(term == "sexM") |> 
    filter(p.value < p_value_cutoff) |> 
    pull(Assay) 
}

proteins_associated_sex_16_24 <- intersect(extract_significant_assays_sex(data_16, proteins_16_24, model_1),
          extract_significant_assays_sex(data_24, proteins_16_24, model_1)) 

extract_significant_assays_bmi <- function(my_data, my_proteins, my_model) {
    my_data  |>
    filter(Assay %in% my_proteins) |> 
    group_by(Assay)  |>  
    do(tidy(lm(my_model, data = .))) |> 
    filter(term == "bmi") |> 
    filter(p.value < p_value_cutoff) |> 
    pull(Assay) 
}

proteins_associated_bmi <- intersect(extract_significant_assays_bmi(data_16, proteins_16_24, model_1),
extract_significant_assays_bmi(data_24, proteins_16_24, model_1))

extract_significant_assays_bodyfat <- function(my_data, my_proteins, my_model) {
    my_data  |>
    filter(Assay %in% my_proteins) |> 
    group_by(Assay)  |>  
    do(tidy(lm(my_model, data = .))) |> 
    filter(term == "bioimp_body_fat") |> 
    filter(p.value < p_value_cutoff) |> 
    pull(Assay) 
}

proteins_associated_bodyfat <- extract_significant_assays_bodyfat(data_24, proteins_24, model_2)


proteins_associated_bodyfat_w_p_value <- data_24  |>
    filter(Assay %in% proteins_24) |> 
    group_by(Assay)  |>  
    do(tidy(lm(model_2, data = .))) |> 
    filter(term == "bioimp_body_fat") |> 
    filter(p.value < p_value_cutoff) |> 
    select(Assay, p.value) |> 
    arrange(p.value)

supptable6  <- proteins_associated_bodyfat_w_p_value |>   
  ungroup() |> 
  mutate(p.value = format(p.value, scientific = TRUE, digits = 2, nsmall = 2)) |>
  left_join(BAMSE_data |> distinct(Assay, Gene_desc), by= "Assay") |>
  mutate(Gene_desc = case_when(Assay == "DEFB4A_DEFB4B" ~ "Defensin beta 4A/B", TRUE ~ Gene_desc)) |> 
  rename("p-value" = p.value, "Protein name" = Gene_desc) |>  
  mutate("Significantly associated with sex after adjustments" = case_when(Assay %in% extract_significant_assays_sex(data_24, proteins_24, model_2) ~ "Yes", TRUE ~ "No")) 

```



\newpage

## Figure 5
```{r}

# Figure 5A
selected_proteins <- proteins_associated_bmi

data_to_plot <- extra_meta_data |> 
  select(subject_id, bmi_z_score_4y, bmi_z_score_8y, bmi_z_score_16y, bmi_z_score_24y) |> 
  pivot_longer(-subject_id, names_to = "follow_up", values_to = "bmi") |> 
  mutate(follow_up = case_when(follow_up == "bmi_z_score_4y" ~ 4,
                               follow_up == "bmi_z_score_8y" ~ 8,
                               follow_up == "bmi_z_score_16y" ~ 16,
                               follow_up == "bmi_z_score_24y" ~ 24) |> factor()) |> 
  left_join(BAMSE_data |> distinct(daid, subject_id, follow_up), by = c("subject_id","follow_up")) |> 
  left_join(BAMSE_data |>
              distinct(daid, Assay, NPX, sex) |> 
              filter(Assay %in% selected_proteins),
            by = c("daid")) |> 
  filter(!is.na(bmi)) |> 
  mutate(sex = as.character(sex)) |> 
  mutate(sex = case_when(sex == "F" ~ "Female", 
                         sex == "M" ~"Male")) |> 
  mutate(follow_up = paste0("Follow-up ", follow_up) |> factor(levels = c("Follow-up 4", "Follow-up 8","Follow-up 16", "Follow-up 24" )))

p1 <- 
  ggplot(data_to_plot, aes(x = bmi, y = NPX, color = sex, fill = sex))+
  geom_point(size = 1, alpha = 0.7, shape = 16) +
  labs(x="BMI (z-scores)", y = "NPX")+
  scale_fill_manual(values=c(col_F, col_M))+
  scale_color_manual(values=c(col_F, col_M))+
  scale_y_continuous(expand = expansion(mult = 0.3))+
  stat_cor(method = "spearman", cor.coef.name = "rho", size = 2.5)+
  facet_grid(rows = vars(Assay), cols = vars(follow_up), scales = "free_y")+
  theme_bw()+
  theme(legend.position = "bottom",
        legend.title = element_blank(), 
        axis.text = element_text(size = 8),
        aspect.ratio = 1)

suppdata$fig5a  <- data_to_plot  |> select(Assay, follow_up,  NPX, bmi, sex)  |> arrange(Assay, follow_up, NPX)

# Figure 5B
selected_proteins <- proteins_associated_bodyfat_w_p_value |> ungroup() |> slice_min(order_by = p.value, n = 8) |> pull(Assay)

data_to_plot <- BAMSE_data |>
  select(daid, subject_id, Assay, NPX, follow_up, sex) |> 
  filter(follow_up %in% c(24)) |> 
  filter(Assay %in% selected_proteins) |>
  left_join(extra_meta_data |> select(subject_id, bioimp_body_fat), by = "subject_id") |> 
  filter(!is.na(bioimp_body_fat)) |> 
  mutate(sex = as.character(sex)) |> 
  mutate(sex = case_when(sex == "F" ~ "Female", 
                         sex == "M" ~"Male"))


p2 <- 
  ggplot(data_to_plot, aes(x = bioimp_body_fat, y = NPX, color = sex, fill = sex))+
  geom_point(size = 1, alpha = 0.7, shape = 16) +
  labs(x="Body fat %", y = "NPX")+
  scale_fill_manual(values=c(col_F, col_M))+
  scale_color_manual(values=c(col_F, col_M))+
  scale_y_continuous(expand = expansion(mult = 0.3))+
  stat_cor(method = "spearman", cor.coef.name = "rho", size = 2.5)+
  facet_wrap(~Assay, ncol = 4, scales = "free")+
  theme_bw()+
  theme(legend.position = "none",
        legend.title = element_blank(), 
        axis.text = element_text(size = 8),
        aspect.ratio = 1)


suppdata$fig5b  <- data_to_plot  |> select(Assay, NPX, bioimp_body_fat, sex)  |> arrange(Assay, NPX)

# Arrange and save Figure 5
p <- ggarrange(p1,p2, nrow = 2, heights = c(3, 2) ,labels = c("a", "b"))

ggsave(paste0("output/Figure 5 ", Sys.Date(),".pdf"), height = 10, width = 8)
```

\newpage
# Supptables

## Supp table 1
```{r}
supptables$Supplementary_table_1 <- BAMSE_data |> 
  distinct(Assay, OlinkID, UniProt) |> 
  left_join(age_signif_proteins |> 
              select(Assay, estimate, p_value, Adjusted_pval, comparison) |>
              mutate(comparison = case_when(comparison == "Follow-up 4 vs 8" ~ "follow-up 4 vs 8",
                                            comparison == "Follow-up 8 vs 16" ~ "follow-up 8 vs 16",
                                            comparison == "Follow-up 16 vs 24" ~ "follow-up 16 vs 24")) |> 
              mutate(Adjusted_pval = format(Adjusted_pval, scientific = TRUE, digits = 2, nsmall = 2)) |> 
              mutate(p_value = format(p_value, scientific = TRUE, digits = 2, nsmall = 2)) |> 
              rename("p-value FDR" = Adjusted_pval, "p-value raw" = p_value, "fold change" = estimate) |> 
              pivot_wider(names_from = comparison, values_from = c("p-value raw", "p-value FDR","fold change"), names_sep = " "), by = "Assay") |> 
    relocate(Assay, UniProt, OlinkID, 
             `p-value raw follow-up 4 vs 8`,`p-value FDR follow-up 4 vs 8`, `fold change follow-up 4 vs 8`, 
             `p-value raw follow-up 8 vs 16`, `p-value FDR follow-up 8 vs 16`,`fold change follow-up 8 vs 16`, 
             `p-value raw follow-up 16 vs 24`,`p-value FDR follow-up 16 vs 24`,`fold change follow-up 16 vs 24`) |> left_join(saved_protein_cluster, by = "Assay")

```

## Supp table 2
```{r}
supptables$Supplementary_table_2 <- supptable2 
```

## Supp table 3
```{r}
supptables$Supplementary_table_3 <- 
  bind_rows(age_signif_proteins_F |> 
              select(Assay, estimate, p_value, Adjusted_pval, comparison) |>
              mutate(comparison = case_when(comparison == "Follow-up 4 vs 8" ~ "females follow-up 4 vs 8",
                                            comparison == "Follow-up 8 vs 16" ~ "females follow-up 8 vs 16",
                                            comparison == "Follow-up 16 vs 24" ~ "females follow-up 16 vs 24")),
            age_signif_proteins_M |> 
              select(Assay, estimate, p_value, Adjusted_pval, comparison) |>
              mutate(comparison = case_when(comparison == "Follow-up 4 vs 8" ~ "males follow-up 4 vs 8",
                                            comparison == "Follow-up 8 vs 16" ~ "males follow-up 8 vs 16",
                                            comparison == "Follow-up 16 vs 24" ~ "males follow-up 16 vs 24"))) |> 
  mutate(Adjusted_pval = format(Adjusted_pval, scientific = TRUE, digits = 2, nsmall = 2)) |> 
  mutate(p_value = format(p_value, scientific = TRUE, digits = 2, nsmall = 2)) |>
  rename("p-value FDR" = Adjusted_pval, "p-value raw" = p_value, "fold change" = estimate) |>
  pivot_wider(names_from = comparison, values_from = c("p-value raw", "p-value FDR","fold change"), names_sep = " ") |> 
  arrange(Assay) |> 
  left_join(BAMSE_data |> distinct(Assay, OlinkID, UniProt), by = "Assay")  |> 
  relocate(Assay, OlinkID, UniProt)

```


## Supp table 4 
```{r}
supptables$Supplementary_table_4 <- BAMSE_data |> 
  distinct(Assay, OlinkID, UniProt) |> 
  left_join(save_p_values_sex_including_raw_p |> 
              select(Assay, estimate, p_value, Adjusted_pval, comparison) |> 
              mutate(comparison = case_when(comparison == "Follow up 4" ~ "follow-up 4",
                                            comparison == "Follow up 8" ~ "follow-up 8",
                                            comparison == "Follow up 16" ~ "follow-up 16",
                                            comparison == "Follow up 24" ~ "follow-up 24")) |> 
              mutate(Adjusted_pval = format(Adjusted_pval, scientific = TRUE, digits = 2, nsmall = 2)) |> 
  mutate(p_value = format(p_value, scientific = TRUE, digits = 2, nsmall = 2)) |> 
              rename("p-value FDR" = Adjusted_pval, "p-value raw" = p_value, "fold change" = estimate) |>
              mutate(comparison = paste0(comparison = paste0("female vs male (",comparison,")" ))) |> 
              pivot_wider(names_from = comparison, values_from = c("p-value raw", "p-value FDR","fold change"), names_sep = " "), by = "Assay") |> 
  
    relocate(Assay, UniProt, OlinkID, 
             `p-value raw female vs male (follow-up 4)`,`p-value FDR female vs male (follow-up 4)`, `fold change female vs male (follow-up 4)`, 
             `p-value raw female vs male (follow-up 8)`,`p-value FDR female vs male (follow-up 8)`, `fold change female vs male (follow-up 8)`, 
             `p-value raw female vs male (follow-up 16)`, `p-value FDR female vs male (follow-up 16)`, `fold change female vs male (follow-up 16)`, 
             `p-value raw female vs male (follow-up 24)`, `p-value FDR female vs male (follow-up 24)`, `fold change female vs male (follow-up 24)`) 

```
## Supp table 5
```{r}
supptables$Supplementary_table_5 <- supptable5
```


## Supp table 6 
```{r}
supptables$Supplementary_table_6 <- supptable6 
```


## Write supplementary tables

```{r}
write_xlsx(supptables, path = paste0("output/Supplementary tables ", Sys.Date(), ".xlsx") )
```

# Write supplementary data
```{r}
write_xlsx(suppdata, path = paste0("output/Supplementary data ", Sys.Date(), ".xlsx") )
```


# Replication analysis of Liu et.al and Niu et.al
```{r}

protein_wide <- BAMSE_data %>%
  dplyr::select(subject_id, follow_up, Assay,  age, sex,NPX) |>#age, sex,
  group_by(subject_id, follow_up, Assay) %>%
  
  pivot_wider(names_from = Assay, values_from = NPX) |>
  arrange(subject_id, follow_up) #%>%  # Order rows by subject_id, follow_up
rownames(protein_wide)<-paste(protein_wide$subject_id, protein_wide$follow_up, sep = "_")



Olink_Data_all<-protein_wide[,-c(1:4)]
rownames(Olink_Data_all)<-rownames(protein_wide)
cov<-protein_wide[,c(1:4)]
rownames(cov)<-rownames(protein_wide)

### inverse normal transformation INT
inormal <- function(x)
{
  qnorm((rank(x, na.last = "keep") - 0.5) / sum(!is.na(x)))
}

### inverse normal transformation INT
Data.H= as.data.frame(apply(t(Olink_Data_all), 1, inormal))


 #Time-course expression analysis of significant proteins from mixed-effect model 
 
tdat<-Olink_Data_all_zscore

# Function to fit mixed model with Kenward-Roger correction
f.RLM.Adjusted.Robust.par <- function(pcol, X) {
  # Validate column index
  if (!(pcol %in% seq_len(ncol(tdat)))) {
    stop(paste("Error: Column index", pcol, "is out of bounds."))
  }
  
  # Ensure `tdat` and `X` have matching rows
  if (nrow(tdat) != nrow(X)) {
    stop("Error: `tdat` and `X` must have the same number of rows!")
  }
  
  # Bind protein data and covariates
  D <- cbind(p = tdat[, pcol], X) %>% as.data.frame()
  
  # Fit models with Kenward–Roger correction
  mod1 <- try(lmer(D[,1] ~ age + sex + (1 | subject_id), data = D), silent = TRUE)
  
  # Define default output structure
  cf <- setNames(rep(NA, 3), c("Age_Estimate", "Age_Std.Error", "Age_p.value"))
  
  if (inherits(mod1, "try-error")) {
    message("Model fitting failed for column: ", pcol)
  } else {
    cft1 <- tryCatch(coef(summary(mod1)), error = function(e) NULL)
    
    # Check if required predictors exist before extracting
    if (!is.null(cft1)) {
      cf[1:3] <- cft1["age", c("Estimate", "Std. Error", "Pr(>|t|)")]
    }
  }
  
  return(cf)
}

# parallel Processing Setup
time0 <- Sys.time()
num_cores <- detectCores() - 1  # Use all but 1 core
c1 <- makePSOCKcluster(num_cores)

# oad required libraries on worker nodes
clusterEvalQ(c1, {
  library(lme4)
  library(lmerTest)
  library(pbkrtest)
  library(dplyr)
})

# xport required variables & function to worker nodes
clusterExport(c1, c('f.RLM.Adjusted.Robust.par', 'cov', 'tdat'))

# run models in parallel across proteins
indiv.results.list <- parLapply(c1, 
                                setNames(seq_len(ncol(tdat)), colnames(tdat)), 
                                function(z) f.RLM.Adjusted.Robust.par(z, cov)
)

# top cluster after computation
stopCluster(c1)

# Compute time taken
Sys.time() - time0

# coerce the list back to a data.frame
indiv.results<- ldply(lapply(indiv.results.list, rbind))
colnames(indiv.results)<-c("Protein","coef", "sd", "P.value")
indiv.results$FDR_BH<-p.adjust(indiv.results$P.value, "BH")
dim(indiv.results[indiv.results$FDR_BH<0.05,])


write.table(indiv.results, file ="Ageassociationmixedeffect.txt",col.names = TRUE, row.names = FALSE,sep = "\t",append=FALSE,quote =FALSE)



#the overlap of  Niu et.al and Bergström et.al. mixed-effect model 


library(ggplot2)
library(irr)  # for Cohen's Kappa
library(readxl)
library(readr)
Liu_Niu_Bergström_overlapping_age_sex <- read_excel("data/Liu_Niu_Bergström_overlapping_age_sex.xlsx", sheet = "Niu 2025 age")
Age.rep<-Ageassociationmixedeffect %>% inner_join(Liu_Niu_Bergström_overlapping_age_sex[,c("Protein","Coef")],by="Protein")
# Your data
agree_df <- data.frame(
  Set1 = Age.rep$Coef,
  Set2_P = Age.rep$coef,
  FDR_BH<-Age.rep$FDR_BH
)

# Agreement in direction
agree_df$Direction <- ifelse(sign(agree_df$Set1) == sign(agree_df$Set2_P), "Same", "Opposite")

# Spearman correlation
cor_res <- cor.test(agree_df$Set1, agree_df$Set2_P, method = "spearman")

rho <- cor_res$estimate
p_val <-format.pval(cor_res$p.value, digits = 3, eps = 1e-16)

cor_text <- paste0(
  "Spearman \u03C1 = ", round(rho, 2),
  "\nP-value = ", p_val
)
# Count observations per group
dir_counts <- table(agree_df$Direction)



agree_df$Direction <- ifelse(sign(agree_df$Set1) == sign(agree_df$Set2_P),
                             "Same", "Opposite")

agree_df$Significance <- ifelse(agree_df$FDR_BH < 0.05,
                                "Significant", "Not significant")

# Create labels with counts
agree_df$ColorGroup <- with(
  agree_df,
  ifelse(
    Significance == "Not significant",
    "Not significant",
    Direction
  )
)

# Fix legend order
agree_df$ColorGroup <- factor(
  agree_df$ColorGroup,
  levels = c("Same", "Opposite", "Not significant")
)



# Count per color group
group_counts <- table(agree_df$ColorGroup)

# Create legend labels with n
legend_labels <- c(
  "Same" = paste0("Same direction (FDR < 0.05, n = ", group_counts["Same"], ")"),
  "Opposite" = paste0("Opposite direction (FDR < 0.05, n = ", group_counts["Opposite"], ")"),
  "Not significant" = paste0("Not significant in  Bergström et al.(n = ", group_counts["Not significant"], ")")
)

# Scatter plot
p <- ggplot(agree_df, aes(x = Set1, y = Set2_P, color = ColorGroup)) +
  geom_point(size = 4) +
  geom_vline(xintercept = 0, linetype = "dotted", color = "black") +
  geom_hline(yintercept = 0, linetype = "dotted", color = "black") +
  annotate(
    "text",
    x = min(agree_df$Set1),
    y = max(agree_df$Set2_P),
    label = cor_text,
    hjust = 0,
    vjust = 1,
    size = 5
  ) +
  labs(
    title = "Consistency of age-associated protein effects between Niu et al. and Bergström et al.",
    x = expression(log[2]~FC~"("~italic(Niu~et~al.)~")"),
    y = expression(log[2]~FC~"("~italic(Bergström~et~al.)~")"),
    color = "Effect direction"
  ) +
  scale_color_manual(
    values = c(
      "Same" = "green",
      "Opposite" = "red",
      "Not significant" = "grey70"
    ),
    labels = legend_labels
  ) +
  theme_minimal()

print(p)

ggsave(
  filename = "agreement_plotwith_Niu.pdf",
  plot = p,
  width = 12,
  height =8,
  units = "in",
  device = cairo_pdf
)



# the overlap of  Liu et.al and Bergström et.al. mixed-effect model 

Liu_GOterms_agepatterns <- read_excel("data/Liu_GOterms_agepatterns.xlsx", 
                                      sheet = "Table S2", skip = 2)
LuiData<-Liu_GOterms_agepatterns[,c(4,148,150,151,152,153,154)]
LuiData<-LuiData[complete.cases(LuiData),]
colnames(LuiData)<-c("Protein","Protein IDs","Con_Slope","Con_Quad","Con_Linear_pvalue","Con_Quad_pvalue","NP_trends" )
LuiData<-LuiData[which(LuiData$Con_Linear_pvalue<0.05),]
Age.rep<-Ageassociationmixedeffect %>% inner_join(LuiData[,c("Protein","Con_Slope")],by="Protein")
# Your data
agree_df <- data.frame(
  Set1 = Age.rep$Con_Slope,
  Set2_P = Age.rep$coef,
  FDR_BH<-Age.rep$FDR_BH
)

# Agreement in direction
agree_df$Direction <- ifelse(sign(agree_df$Set1) == sign(agree_df$Set2_P), "Same", "Opposite")

# Spearman correlation
cor_res <- cor.test(agree_df$Set1, agree_df$Set2_P, method = "spearman")

rho <- cor_res$estimate
p_val <-format.pval(cor_res$p.value, digits = 3, eps = 1e-16)

cor_text <- paste0(
  "Spearman \u03C1 = ", round(rho, 2),
  "\nP-value = ", p_val
)
# Count observations per group
dir_counts <- table(agree_df$Direction)


agree_df$Direction <- ifelse(sign(agree_df$Set1) == sign(agree_df$Set2_P),
                             "Same", "Opposite")

agree_df$Significance <- ifelse(agree_df$FDR_BH < 0.05,
                                "Significant", "Not significant")


# Create labels with counts
agree_df$ColorGroup <- with(
  agree_df,
  ifelse(
    Significance == "Not significant",
    "Not significant",
    Direction
  )
)

# Fix legend order
agree_df$ColorGroup <- factor(
  agree_df$ColorGroup,
  levels = c("Same", "Opposite", "Not significant")
)


# Count per color group
group_counts <- table(agree_df$ColorGroup)

# Create legend labels with n
legend_labels <- c(
  "Same" = paste0("Same direction (FDR < 0.05, n = ", group_counts["Same"], ")"),
  "Opposite" = paste0("Opposite direction (FDR < 0.05, n = ", group_counts["Opposite"], ")"),
  "Not significant" = paste0("Not significant in  Bergström et al.(n = ", group_counts["Not significant"], ")")
)

# Scatter plot
p <- ggplot(agree_df, aes(x = Set1, y = Set2_P, color = ColorGroup)) +
  geom_point(size = 4) +
  geom_vline(xintercept = 0, linetype = "dotted", color = "black") +
  geom_hline(yintercept = 0, linetype = "dotted", color = "black") +
  annotate(
    "text",
    x = min(agree_df$Set1),
    y = max(agree_df$Set2_P),
    label = cor_text,
    hjust = 0,
    vjust = 1,
    size = 5
  ) +
  labs(
    title = "Consistency of age-associated protein effects between Liu et al. and Bergström et al.",
    x = expression(log[2]~FC~"("~italic(Liu~et~al.)~")"),
    y = expression(log[2]~FC~"("~italic(Bergström~et~al.)~")"),
    color = "Effect direction"
  ) +
  scale_color_manual(
    values = c(
      "Same" = "green",
      "Opposite" = "red",
      "Not significant" = "grey70"
    ),
    labels = legend_labels
  ) +
  theme_minimal()

print(p)

ggsave(
  filename = "agreement_plotwith_Niu.pdf",
  plot = p,
  width = 12,
  height =8,
  units = "in",
  device = cairo_pdf
)




```





